Supp mat: time series paper

### 2017-08-16 writing/manuscript_draft.Rmd
updated methods section

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. We only briefly discuss this approach as it has been described elsewhere. Essentially, we use a population model and simulate it for a number of years repetitively. We can then determine the minimum time series required to achieve designed levels of significance and power for a particular question. This approach requires us to determine values for our model parameters (e.g. population variability). As an example, we can take the following population model for population size $N$ at time $t$:
\begin{equation}
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase). It is important to note that any population model could be subsituted for equation \ref{sim_model}, as we do in the supplementary material (Figs. \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).

We can determine the minimum time series required by simulating this model repetitively for different lengths of time. Here, statistical power is simply the fraction of simulations with significant trends in abundance. A significant trend in abundance is determined by estimating the significance of the slope parameter from simple linear regression. Then, the minimum time required ($T_{min}$) is the number of years required to obtain statistical power above some pre-determined threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015]. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The database includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The database also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations in the database ($n =$ `r nrow(LPI_pop_info)`), we were able to connect a database on biological characteristics [@Myhrvold2015], including body size and generation length. All `r nrow(LPI_pop_info)` populations are within the Aves class. In the supplementary material, we explore how the minimum time required does not strongly correlate with any biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 40 years would have 39 possible contiguous subsamples of length 2, 38 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 40 [@Gerber1999; @Giron-Nava2017]. Then, we determine the $T_{min}$ required to be confident (i.e. have high enough statistical power) in an estimate of the slope coefficient from linear regression. Therefore, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero given the long-term, or ``true'', time series also has a significant slope. This is a measure of statistical power (often denoted as $1 - \beta$, where $\beta$ is the probability of a type II error). Lastly, we determine which subsample length is required to guarantee we have above a certain threshold of statistical power, which we will assume is 0.8 by convention [@Cohen1992].

>>>>

One approach to determining the minimum time series length needed is through repetitive simulations of a population model [@Gerrodette1987]. This is the same approach one might use in sample size calculations for any experimental design too complicated for simple power analyses [@Bolker2008;@Johnson2015]. We only briefly discuss this approach as it has been described elsewhere. Essentially, we use a population model and repetitively simulate it for a number of years. This approach requires us to determine values for our model parameters (e.g. population variability). As an example, we can take the following population model for population size $N$ at time $t$:

\begin{equation}
N(t + 1) = N(t) + r(t) \mbox{ with } r(t) \sim N(\mu, \sigma)
\label{sim_model}
\end{equation}
where $\epsilon$ is a normally-distributed random noise term with mean $\mu$ and standard deviation $\sigma$. The rate of growth $r$ is also the trend strength of the increase or decrease (i.e. the rate of increase). It is important to note that any population model could be subsituted for equation \ref{sim_model}, as we do in the supplementary material (Figs. \ref{fig:min_time_dist_log_pop}, \ref{fig:shark_example}).

Statistical power is then the fraction of simulations that meet some criteria. Here, we use whether the slope parameter from linear regression is significant. Statistical power changes with the simulation length (Fig. \ref{fig:theoretical_approach}b). Then, the minimum time required ($T_{min}$) is the number of years needed to obtain statistical power above some threshold (historically at 0.8, @Cohen1992). Statistical power of 0.8 would indicate that, if there was a true trend in abundance, we would have a 0.8 probability of detecting the trend. We also tested the effect of varying both the significance level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

## Data source
We use a database of `r length(table(dat$ID))` population time series compiled in [@Keith2015] to compare the predictability of growth rates amongst populations. The data are originally from the Global Population Dynamics Database [@GPDD2010]. We filtered out short time series (less than `r pop_info$series_length[1]` years), and those with missing data, and were left with `r nrow(pop_info)` time series. The data includes information on `r length(table(pop_info$Binomial))` vertebrate species with a focus on mammals, birds, and fish. The data also includes information on generation length and census specifications. For each time series, we also calculated variables of interest like variance in population size, long-term trend in abundance (slope coefficient from simple linear regression), and temporal autocorrelation. <!--Lastly, we also calculated the minimum number of years required (10 years or 3 generations) under IUCN Red List Criterion A2 to classify a species as vulnerable [@IUCN2012].-->

```{r,echo=F}
# Pull in results of biological characteristics comparison to min time series length required
load('../cleaned-data/combined_databases_with_min_time_linear_regression.Rdata')
```
For a subset of populations ($n =$ `r nrow(LPI_pop_info)`), we had information on biological characteristics from another paper [@Myhrvold2015], including body size and generation length. All `r nrow(LPI_pop_info)` populations were birds. We examine how the minimum time required is related to these biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to include all necessary information (e.g. variability) about the population. In other words, each time series is a representative sample. We first take all possible contiguous subsamples of each time series. For example, a time series of 35 years would have 34 possible contiguous subsamples of length 2, 34 possible contiguous subsamples of length 3, and continuing until 1 possible contiguous subsample of length 35 [@Gerber1999; @Giron-Nava2017]. Next, we run a linear regression for each subsampled time series. Then, we determine the fraction of subsamples of a particular length that have estimated slope coefficients which are statistically different from zero. We only look at the fraction of samples where the long-term, or ``true'', time series also has a significant slope. This fraction is a measure of statistical power. Lastly, we determine which subsample length is required to achieve a certain threshold of statistical power (0.8, @Cohen1992). The subsampled length meeting these criteria is the minimum time series length required ($T_{min}$).

### 2017-08-16 writing/manuscript_draft.tex
updated methods section

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
approach as it has been described elsewhere. Essentially, we use a
population model and simulate it for a number of years repetitively. We
can then determine the minimum time series required to achieve designed
levels of significance and power for a particular question. This
approach requires us to determine values for our model parameters
(e.g.~population variability). As an example, we can take the following
population model for population size \(N\) at time \(t\):

@@

We can determine the minimum time series required by simulating this
model repetitively for different lengths of time. Here, statistical
power is simply the fraction of simulations with significant trends in
abundance. A significant trend in abundance is determined by estimating
the significance of the slope parameter from simple linear regression.
Then, the minimum time required (\(T_{min}\)) is the number of years
required to obtain statistical power above some pre-determined threshold
(historically at 0.8, Cohen (1992)). Statistical power of 0.8 would
indicate that, if there was a true trend in abundance, we would have a
0.8 probability of detecting the trend. We also tested the effect of
varying both the significance level and statistical power (Fig.
\ref{fig:min_time_vs_alpha_beta}).

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015). The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The database includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The
database also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation.

For a subset of populations in the database (\(n =\) 547), we were able
to connect a database on biological characteristics (Myhrvold et al.
2015), including body size and generation length. All 547 populations
are within the Aves class. In the supplementary material, we explore how
the minimum time required does not strongly correlate with any
biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 40 years would have 39 possible contiguous
subsamples of length 2, 38 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 40
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Then, we
determine the \(T_{min}\) required to be confident (i.e.~have high
enough statistical power) in an estimate of the slope coefficient from
linear regression. Therefore, we run a linear regression for each
subsampled time series. Then, we determine the fraction of subsamples of
a particular length that have estimated slope coefficients which are
statistically different from zero given the long-term, or ``true'', time
series also has a significant slope. This is a measure of statistical
power (often denoted as \(1 - \beta\), where \(\beta\) is the
probability of a type II error). Lastly, we determine which subsample
length is required to guarantee we have above a certain threshold of
statistical power, which we will assume is 0.8 by convention (Cohen
1992).

>>>>

One approach to determining the minimum time series length needed is
through repetitive simulations of a population model (Gerrodette 1987).
This is the same approach one might use in sample size calculations for
any experimental design too complicated for simple power analyses
(Bolker 2008; P. C. Johnson et al. 2015). We only briefly discuss this
approach as it has been described elsewhere. Essentially, we use a
population model and repetitively simulate it for a number of years.
This approach requires us to determine values for our model parameters
(e.g.~population variability). As an example, we can take the following
population model for population size \(N\) at time \(t\):

@@

Statistical power is then the fraction of simulations that meet some
criteria. Here, we use whether the slope parameter from linear
regression is significant. Statistical power changes with the simulation
length (Fig. \ref{fig:theoretical_approach}b). Then, the minimum time
required (\(T_{min}\)) is the number of years needed to obtain
statistical power above some threshold (historically at 0.8, Cohen
(1992)). Statistical power of 0.8 would indicate that, if there was a
true trend in abundance, we would have a 0.8 probability of detecting
the trend. We also tested the effect of varying both the significance
level and statistical power (Fig. \ref{fig:min_time_vs_alpha_beta}).

@@

We use a database of 2444 population time series compiled in (Keith et
al. 2015) to compare the predictability of growth rates amongst
populations. The data are originally from the Global Population Dynamics
Database (NERC Centre for Population Biology 2010). We filtered out
short time series (less than 35 years), and those with missing data, and
were left with 878 time series. The data includes information on 478
vertebrate species with a focus on mammals, birds, and fish. The data
also includes information on generation length and census
specifications. For each time series, we also calculated variables of
interest like variance in population size, long-term trend in abundance
(slope coefficient from simple linear regression), and temporal
autocorrelation.

For a subset of populations (\(n =\) 547), we had information on
biological characteristics from another paper (Myhrvold et al. 2015),
including body size and generation length. All 547 populations were
birds. We examine how the minimum time required is related to these
biological characteristics (Fig. \ref{fig:biological_correlates}).

@@

We assume that each time series in our database is long enough to
include all necessary information (e.g.~variability) about the
population. In other words, each time series is a representative sample.
We first take all possible contiguous subsamples of each time series.
For example, a time series of 35 years would have 34 possible contiguous
subsamples of length 2, 34 possible contiguous subsamples of length 3,
and continuing until 1 possible contiguous subsample of length 35
(Gerber, DeMaster, and Kareiva 1999; Giron-Nava et al. 2017). Next, we
run a linear regression for each subsampled time series. Then, we
determine the fraction of subsamples of a particular length that have
estimated slope coefficients which are statistically different from
zero. We only look at the fraction of samples where the long-term, or
``true'', time series also has a significant slope. This fraction is a
measure of statistical power. Lastly, we determine which subsample
length is required to achieve a certain threshold of statistical power
(0.8, Cohen (1992)). The subsampled length meeting these criteria is the
minimum time series length required (\(T_{min}\)).
